---
title: "stipm AR1 sim"
author: "dan.auerbach@dfw.wa.gov modifying DeFilippo et al. 2021 AR1 model and fitting to simulated data in consultation with Thomas.Buehrens@dfw.wa.gov"
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
output: 
  wdfwTemplates::wdfw_html_format
---

This script fits the AR1 model [original implementation](https://github.com/lukasdefilippo/ST-IPM) from [DeFilippo et al.'s (2021)](https://www.sciencedirect.com/science/article/pii/S0165783621001429) "Spatiotemporal Integrated Population Model" and tests alternative parameterizations to improve convergence on simulated data.

# Setup
```{r setup, results = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, results = FALSE, warning = FALSE, message = FALSE)
library("tidyverse")
library("gt")
library("odbc"); library("DBI")
library("rstan")
library("bayesplot")
options(mc.cores = 10) #12 on DA machine
rstan_options(auto_write = TRUE)
theme_set(theme_light())
```

# Function to simulate AR1 dataset
```{r}
make_sim_data<-function(yrs, pops){
   #data
  n_year = yrs#14
  n_pop = pops#34
  n_pop_tot = pops#34
  pop_tot = 1:n_pop
  n_tot = (n_year-1) * n_pop
  
  
  #params
  sigma_obs = rlnorm(n_pop,-1,0.5)
  sigma_proc = rlnorm(n_pop,-1,0.5)
  mu = rnorm(n_pop, 8, 2)
  phi = rbeta(n_pop,3,3)
  init = rnorm(n_pop,mu,sqrt(sigma_proc^2/(1-phi^2)))
  log_adult_est = matrix(NA,n_year,n_pop)
  adult_obs= matrix(NA,n_year-1,n_pop)
  proc_dev = matrix(NA,n_year-1,n_pop)
  for(i in 1:n_pop){
  	for(y in 1:n_year){
  		if(y==1){
  			log_adult_est[y,i] = init[i];
  			adult_obs[y,i] = rlnorm(1,log_adult_est[y,i],sigma_obs[i])
  		}else{
  		  proc_dev[y-1, i] = rnorm(1,0,sigma_proc[i])
  			log_adult_est[y,i] = mu[i] + phi[i]*(log_adult_est[y-1,i]-mu[i]) + proc_dev[y-1, i];
  			if(y<n_year){
  			  adult_obs[y,i] = rlnorm(1,log_adult_est[y,i],sigma_obs[i])
  			}
  		}
  	}
  }
  adult_obs_long<-as_tibble(adult_obs)%>%
    rowid_to_column()%>%
    rename(year_id=rowid)%>%
    pivot_longer(cols = !"year_id")%>%
    mutate(pop_id=as.integer(str_replace(name,"V","")))%>%
    rename(tot_dat=value)%>%
    select(-name)%>%
    arrange(pop_id,year_id)
  
  #more data
  tot_dat = adult_obs_long$tot_dat
  tot_dat_yr = adult_obs_long$year_id
  tot_dat_pop = adult_obs_long$pop_id
  tot_true = adult_obs_long$year_id
  slice_tot_start=adult_obs_long%>%
    rownames_to_column()%>%
    group_by(pop_id)%>%
    filter(year_id==1)%>%
    ungroup()%>%
    select(rowname)%>%
    mutate(rowname = as.integer(rowname))%>%
    deframe()
  slice_tot_end<-adult_obs_long%>%
    rownames_to_column()%>%
    group_by(pop_id)%>%
    filter(year_id==max(year_id))%>%
    ungroup()%>%
    select(rowname)%>%
    mutate(rowname = as.integer(rowname))%>%
    deframe()
  
  stan_data = list(
    n_year = n_year,
    n_pop = n_pop,
    n_pop_tot = n_pop_tot,
    pop_tot = pop_tot,
    n_tot = n_tot,
    tot_dat = tot_dat,
    tot_true = tot_true,
    slice_tot_start = slice_tot_start,
    slice_tot_end = slice_tot_end,
    tot_dat_yr = tot_dat_yr,
    tot_dat_pop = tot_dat_pop
  )
  
  params = list(
    sigma_obs = sigma_obs,
    sigma_proc = sigma_proc,
    mu = mu,
    phi = phi,
    init = init,
    log_adult_est = log_adult_est,
    adult_obs= adult_obs,
    proc_dev = proc_dev
  )
  results <- list(stan_data,params)
  names(results)[[1]]<-"stan_data"
  names(results)[[2]]<-"params"
  return(results)
}
```

Next, wrappers to the `rstan::stan` function facilitate repeated fitting with consistent control parameters and input data lists.

```{r ar1_functions}
stan_ar1_sim <- function(stan_data,
                         modelfile, 
                         n_iter = 2000, 
                         n_chain = 2, 
                         init="random",
                         warmup = 1000,
                         thin = 1,
                         control = list(adapt_delta = 0.99, max_treedepth = 10.25)
                         )
  {
  #note following orig naming convention where "tot" refers to "total return" = spwn + hvst = rtrn, as estimated from FRAM
  stan_fit <- stan(
    file = modelfile,
    iter = n_iter, 
    chains = n_chain, 
    seed = 222,
    init=init,
    warmup = warmup,
    thin = thin,
    control = control,
    data = stan_data
  )
  return(stan_fit)
}
```

# Simulate AR1 data, fit AR1 model to simulated AR1 data
```{r}
#simulate data
simdat<-make_sim_data(yrs = 40, pops = 2)


#fit model
fit_ar1 <- stan_ar1_sim(stan_data = simdat$stan_data,
                        modelfile = 
                        #'LD_coho_forecast_AR_ind_2.stan', #lukas original
                        'LD_coho_forecast_AR_ind_2_v2.stan', #thomas reparam but preserve data format
                        #'LD_coho_forecast_AR_ind_2_v3.stan', #thomas reparam, new data format
                        n_iter = 2000,
                        n_chain = 4,
                        warmup = 1000,
                        thin = 1,
                        control = list(adapt_delta = 0.995, max_treedepth = 11)
                        )

#fit diagnostics
# posterior<-as.array(fit_ar1)
# np<-nuts_params(fit_ar1)
# color_scheme_set("darkgray")
# mcmc_pairs(posterior, np = np, pars = c("mu[1]","sigma_proc[1]","sigma_obs[1]","phi[1]","init[1]"),
#            off_diag_args = list(size = 0.75))

#save model summary
write.csv(summary(fit_ar1)$summary,"result.csv")

#print "true" parameters from simulation
print(simdat$params)

#print model estimated posterior medians to compare with "truth"
summary(fit_ar1)$summary%>%
  as.data.frame()%>%
  rownames_to_column()%>%
  as_tibble()%>%
  select(rowname,`50%`)%>%
  print(n=200)
```


Try Eric Ward AR1
```{r}
library(devtools)
install_github("nwfsc-timeseries/atsar")
library(atsar)

plot(simdat$params$adult_obs[,1],type="b")

ss_ar = fit_stan(y = simdat$params$adult_obs[,1],
                 est_drift=FALSE,
                 model_name = "ss_ar",
                 family = "lognormal",
                 mcmc_list=list(
                   n_mcmc = 2000,
                   n_burn = 1000,
                   n_chain = 4,
                   n_thin = 1,
                   control = list(adapt_delta=0.9999)
                 )
)
write.csv(summary(ss_ar)$summary,"results_ward.csv")

```